Data migration under a Changing Schema

### 2023-03-13 2022 Migration/articleMigration.tex
WIP

   The information system enforces the semantics by ensuring that the following requirements remain true at all times:
\begin{eqnarray}
   \triple{a}{\declare{n}{A}{B}}{b}\in\triples&\Rightarrow&\declare{n}{A}{B}\in\rels
   \label{eqn:define R}\\
   \forall u\in\rules&:&\sat{u}{\dataset}
   \label{eqn:satisfaction}
\end{eqnarray}
   Requirement~\ref{eqn:define R} ensures that the information system only works with triples whose relation is defined in the schema.
   Requirement~\ref{eqn:satisfaction} ensures that the dataset satisfies all rules in the schema.

   The maintainance relation between roles and rules is used to determine who is allowed to change the dataset.
   For every rule $u$ in the schema, there is a function $\id{viol}_u : \Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$
   that represents the set of violations%

   The rules in the schema serve as invariants,
   i.e.\ constraints that the data in the database must satisfy at all times.
   At times in which the dataset does not satisfy a rule, that rule is said to be {\em broken}.
   In those cases, something or someone must take action to fix the situation and restore invariance of that rule.

   If a triple $\triple{a}{r}{b}$ satisfies requirement~\ref{eqn:define R} and its relation $r$ satisfies requirement~\ref{eqn:relationsIntroduceConcepts},
   this triple is said to {\em befit} schema $\quadruple{\concepts}{\isa}{\rels}{\rules}$.

   The system is {\em at rest} when every rule in the schema is satisfied:
\begin{equation}
\end{equation}
   However, the system continually registers events that insert or delete triples in its dataset,
   which may cause rules to be broken.
   This requires a reaction to restore invariance.
   Example: if a rule says that every request must be acknowledged, the receipt of a request (which is an event) will violate this rule.
   Sending a message that acknowledges this receipt (another event) will cause the rule to be satisfied again.
   During the time between the receipt of the request and sending the acknowledgment, the rule is (temporarily) not satisfied.

   This paper assumes two ways of enforcing satisfaction of rules: automatic or manual.
   Manual enforcement is specified by the developer by assigning a rule $u$ to a role $o$.
   Any actor with role $o$, whether person or machine,
   can restore invariance of rule $u$ by inserting or deleting the right triples.
   For this reason, every rule is assigned to at least one role by means of the relation $\maintain$.
\begin{equation}
   \forall u\in\rules\ \exists o\in\roles:\ o\maintain u\label{eqn:maintain}
\end{equation}
   $o\maintain u$ means that role $o$ must keep rule $u$ satisfied, i.e.\ to keep $\sat{u}{\dataset}$ true.
   As long as rule $u$ is broken, the system should notify actors in the role(s) that are administered in $\maintain$ to do something about it.

@@

   \subsection{Changing data}
   \label{sct:Changing data}
   @Bas, In het onderstaande heb ik mij er nog wat te makkelijk vanaf gemaakt.
   Aandacht nodig.

      Events can cause data to change.
      So, we model events as functions $\Dataset\rightarrow\Dataset$.
      For triples, we distinguish just two types of event: insert and delete events.
      An event is a label (ins or del) with a dataset that is to be inserted in or deleted from another dataset.
      We denote events as $\ins{\pair{\triples_{\id{ev}}}{\inst_{\id{ev}}}}$ and $\del{\pair{\triples_{\id{ev}}}{\inst_{\id{ev}}}}$.
      They are defined by:
   \begin{eqnarray}
      \ins{\pair{\triples_{\id{ev}}}{\inst_{\id{ev}}}}(\dataset)&=&\pair{\triples_{\dataset}\cup\triples_{\id{ev}}}{\inst_{\dataset}\cup\inst_{\id{ev}}}\\
      \del{\pair{\triples_{\id{ev}}}{\inst_{\id{ev}}}}(\dataset)&=&\pair{\triples_{\dataset}-\triples_{\id{ev}}}{\inst_{\dataset}-\inst_{\id{ev}}}
   \end{eqnarray}
      Insert events are represented as $\ins{\dataset}$ and delete events by $\del{\dataset}$.
      Both are functions $\Dataset\rightarrow\Dataset$, which are defined as:
      Since datasets are sets, we will use set union ($\cup$) and set difference ($-$) to make changes in datasets.

@@

The ${\tt CLASSIFY\ }(C',1){\tt\ IS\ }(C,0)$ statement adds two pairs to the $\isa$ relation:
$\pair{(C',1)}{(C,0)}$ and $\pair{(C,0)}{(C',1)}$.
Note that if $\dataset$ is a dataset, and $\dataset'$ is that dataset with pairs added%
\footnote{TODO: het woord `add' is informeel, dit kan beter} to the $\isa$ relation, then the result is again a dataset.
The ${\tt ENFORCE\ }r{\tt\ :=\ }e$ statements add triples $(x,r,y)$ for all $(x,y)\in e$.
The expressions $e$ in these statements are such that $x\ (\inst \compose \kleenestar{\flip{\isa}})\ (A',1)$ and $y \ (\inst \compose \kleenestar{\flip{\isa}})\ (B',1)$, thus preserving the property that the result is a dataset.

>>>>

   The information system enforces the semantics by ensuring that the following requirements remain true at all times:
\begin{eqnarray}
   \triple{a}{\declare{n}{A}{B}}{b}\in\triples&\Rightarrow&\declare{n}{A}{B}\in\rels
   \label{eqn:define R}\\
   \forall u\in\rules&:&\sat{u}{\dataset}
   \label{eqn:satisfaction}
\end{eqnarray}
   Requirement~\ref{eqn:define R} ensures that the information system only works with triples whose relation is defined in the schema.
   Requirement~\ref{eqn:satisfaction} ensures that the dataset satisfies all rules in the schema.
   The rules in the schema serve as invariants,
   i.e.\ constraints that the data in the database must satisfy at all times.
   At times in which the dataset does not satisfy a rule, that rule is said to be {\em broken}.
   In those cases, something or someone must take action to fix the situation and restore invariance of that rule.
   The system is {\em at rest} when every rule in the schema is satisfied.

   % The maintainance relation between roles and rules is used to determine who is allowed to change the dataset.
   % For every rule $u$ in the schema, there is a function $\id{viol}_u : \Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$
   % that represents the set of violations.%

\subsection{Changing data}
\label{sct:Changing data}
   Events can cause a dataset to change.
   So, we model events as functions $\Dataset\rightarrow\Dataset$.
   We will use variable $e$, $e'$, $e''$, and so on to represent events.
   We distinguish just two types of event: insert and delete events.
   An event is a label (\id{ins} or \id{del}) with a dataset that is to be inserted or deleted.
   If $\triples$ is the set of triples inserted or deleted by an event $e$,
   we denote that event either by $\ins{\triples}$ or $\del{\triples}$.
   They are defined by:
\begin{eqnarray}
   e(\dataset)&=&\id{case}\ e\ \id{of}\\
   &&\begin{array}[t]{@{}rcl}
      \ins{\triples}&=&\pair{\triples_{\dataset}\cup\triples}{\id{insInst}_{\triples}(\inst_{\dataset})}\\
      \del{\triples}&=&\pair{\triples_{\dataset}-\triples}{\id{delInst}_{\triples}(\inst_{\dataset})}
   \end{array}
\end{eqnarray}
   in which $\id{insInst}_{\triples}$ transforms $\inst_{\dataset}$ such that $\ins{\triples}(\dataset)$ complies with
   equations~\ref{eqn:specialization} and~\ref{eqn:wellTypedEdge},
   and likewise for $\id{delInst}_{\triples}$.

   As the system registers events, it inserts or deletes triples in its dataset.
   This may violate requirement~\ref{eqn:satisfaction} when one or more rules are broken,
   i.e.\ if there is a rule $u$ for which $\sat{u}{e(\dataset)}$ does not hold.
   In that case, requirement~\ref{eqn:satisfaction} prohibits the system to be in state $e(\dataset)$,
   so either the system stays in the state $\dataset$ or there is a reaction $e'$ that restores invariance of the broken rules:
\begin{equation}
   \sat{u}{\dataset}\Rightarrow\sat{u}{e'(e(\dataset))}
   \label{eqn:restoring}
\end{equation}
   The combination of events $e$ and $e'$ is called a {\em transaction}.

   In practice, there are different ways of enforcing satisfaction of rules.
   One way is to let a user or a computer ``invent'' $e'$.
   A developer can specify this by assigning a rule $u$ to a role $o$.
   Any actor with role $o$
   can restore invariance of rule $u$ by inserting or deleting the right triples.
   For this reason, every rule is assigned to at least one role by means of the relation $\maintain$.
\begin{equation}
   \forall u\in\rules\ \exists o\in\roles:\ o\maintain u\label{eqn:maintain}
\end{equation}
   $o\maintain u$ means that role $o$ must keep rule $u$ satisfied, i.e.\ to keep $\sat{u}{\dataset}$ true.
   As long as rule $u$ is broken, the system should notify actors in the role(s) that are administered in $\maintain$ to do something about it.

   If $\maintain$ specifies no role for a given rule $u$, the system can prevent $e$ from having any effect.
   This implements blocking behaviour, and must be accompanied by an error message.

@@

@@

The ${\tt CLASSIFY\ }(C',1){\tt\ IS\ }(C,0)$ statement adds two pairs to the $\isa$ relation:
$\pair{(C',1)}{(C,0)}$ and $\pair{(C,0)}{(C',1)}$.
Note that if $\dataset$ is a dataset, and $\dataset'$ is that dataset with pairs added%
\footnote{TODO: het woord `add' is informeel, dit kan beter} to the $\isa$ relation, then the result is again a dataset.
The ${\tt ENFORCE\ }r{\tt\ :=\ }e$ statements add triples $(x,r,y)$ for all $(x,y)\in e$.
The terms $t$ in these statements are such that $x\ (\inst \compose \kleenestar{\flip{\isa}})\ (A',1)$ and $y \ (\inst \compose \kleenestar{\flip{\isa}})\ (B',1)$, thus preserving the property that the result is a dataset.
